---
title: "Glasmacinal Microbiome: Original vs RasperGade16S Branch Comparison"
subtitle: "Assessing the effect of 16S rRNA gene copy-number correction on biological conclusions"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo       = TRUE,
    warning    = T,
    message    = T,
    fig.align  = "center",
    dpi        = 300
)

# ── Required packages ─────────────────────────────────────────────────────────
pkgs <- c(
    "tidyverse", "qiime2R", "vegan", "ggplot2", "patchwork", "cowplot",
    "UpSetR", "scales", "ggrepel", "pheatmap", "RColorBrewer",
    "knitr", "kableExtra", "gt", "gtsummary", "ape"
)

for (p in pkgs) {
    if (!requireNamespace(p, quietly = TRUE)) {
        message("Installing: ", p)
        if (p %in% c("qiime2R")) {
            if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
            remotes::install_github("jbisanz/qiime2R")
        } else {
            install.packages(p, quiet = TRUE)
        }
    }
    suppressPackageStartupMessages(library(p, character.only = TRUE))
}
```

```{r}
# =============================================
# Run the preparation script first
# =============================================
cat("=== Running render_comparison.R ===\n\n")

source("render_comparison.R")

cat("\n=== File preparation completed ===\n\n")
```


```{r directories}
rm(list=ls())
# Set this to your project base directory
PROJECT_DIR <- "~/projects/microbiome"
# Source files from workflow output
WORKFLOW_DIR <- file.path(PROJECT_DIR, "workflow_results")
ANALYSIS_DIR <- file.path(WORKFLOW_DIR, "branch_comparison")

setwd(ANALYSIS_DIR)
# ── Directory structure ────────────────────────────────────────────────────────
# Adjust INPUT_DIR to your workflow_results path
INPUT_ORIG  <- file.path(ANALYSIS_DIR,"input/original")       # Branch A artifacts
INPUT_RASP  <- file.path(ANALYSIS_DIR,"input/raspergade")    # Branch B artifacts
INPUT_STATS <- file.path(ANALYSIS_DIR,"input/stats")         # CSV statistical outputs (both branches)

dir.create("figures",       showWarnings = FALSE, recursive = TRUE)
dir.create("tables",        showWarnings = FALSE, recursive = TRUE)
dir.create("results",       showWarnings = FALSE, recursive = TRUE)
dir.create("figures/alpha", showWarnings = FALSE)
dir.create("figures/beta",  showWarnings = FALSE)
dir.create("figures/da",    showWarnings = FALSE)
dir.create("figures/table", showWarnings = FALSE)

# ── Expected input files ───────────────────────────────────────────────────────
# Adjust paths to match your actual directory structure.
# Default: mirrors workflow_revision.sh output directories.
paths <- list(
    # Original branch
    orig_table     = file.path(INPUT_ORIG, "table.qza"),
    orig_taxonomy  = file.path(INPUT_ORIG, "taxonomy.qza"),
    orig_faith     = file.path(INPUT_ORIG, "faith_pd_vector.qza"),
    orig_evenness  = file.path(INPUT_ORIG, "evenness_vector.qza"),
    orig_wunifrac  = file.path(INPUT_ORIG, "weighted_unifrac_distance_matrix.qza"),
    orig_wunifrac_pcoa  = file.path(INPUT_ORIG, "weighted_unifrac_pcoa_results.qza"),

    # RasperGade branch
    rasp_table     = file.path(INPUT_RASP, "table-corrected.qza"),
    rasp_taxonomy  = file.path(INPUT_ORIG, "taxonomy.qza"),  # shared
    rasp_faith     = file.path(INPUT_RASP, "faith_pd_vector.qza"),
    rasp_evenness  = file.path(INPUT_RASP, "evenness_vector.qza"),
    rasp_wunifrac  = file.path(INPUT_RASP, "weighted_unifrac_distance_matrix.qza"),
    rasp_wunifrac_pcoa  = file.path(INPUT_RASP, "weighted_unifrac_pcoa_results.qza"),

    # Metadata
    metadata       = file.path(ANALYSIS_DIR,"input/meta_data.tsv"),

    # CSV stats — original
    orig_faith_anova      = file.path(INPUT_STATS, "original/faith_pd_lmm_anova.csv"),
    orig_faith_emmeans    = file.path(INPUT_STATS, "original/faith_pd_lmm_pairwise_emmeans.csv"),
    orig_even_anova       = file.path(INPUT_STATS, "original/pielou_evenness_lmm_anova.csv"),
    orig_even_emmeans     = file.path(INPUT_STATS, "original/pielou_evenness_lmm_pairwise_emmeans.csv"),
    orig_perm             = file.path(INPUT_STATS, "original/weighted_unifrac_pairwise_adonis.csv"),
    orig_linDA_asv        = file.path(INPUT_STATS, "original/linDA_ASVs_Sig_LFC1.csv"),
    orig_linDA_genus      = file.path(INPUT_STATS, "original/linDA_Genus_Sig_LFC1.csv"),
    orig_linDA_family     = file.path(INPUT_STATS, "original/linDA_Family_Sig_LFC1.csv"),
    orig_linDA_order      = file.path(INPUT_STATS, "original/linDA_Order_Sig_LFC1.csv"),
    orig_linDA_class      = file.path(INPUT_STATS, "original/linDA_Class_Sig_LFC1.csv"),
    orig_linDA_phylum     = file.path(INPUT_STATS, "original/linDA_Phylum_Sig_LFC1.csv"),

    # CSV stats — RasperGade
    rasp_faith_anova      = file.path(INPUT_STATS, "raspergade/faith_pd_lmm_anova.csv"),
    rasp_faith_emmeans    = file.path(INPUT_STATS, "raspergade/faith_pd_lmm_pairwise_emmeans.csv"),
    rasp_even_anova       = file.path(INPUT_STATS, "raspergade/pielou_evenness_lmm_anova.csv"),
    rasp_even_emmeans     = file.path(INPUT_STATS, "raspergade/pielou_evenness_lmm_pairwise_emmeans.csv"),
    rasp_perm             = file.path(INPUT_STATS, "raspergade/weighted_unifrac_pairwise_adonis.csv"),
    rasp_linDA_asv        = file.path(INPUT_STATS, "raspergade/linDA_ASVs_Sig_LFC1.csv"),
    rasp_linDA_genus      = file.path(INPUT_STATS, "raspergade/linDA_Genus_Sig_LFC1.csv"),
    rasp_linDA_family     = file.path(INPUT_STATS, "raspergade/linDA_Family_Sig_LFC1.csv"),
    rasp_linDA_order      = file.path(INPUT_STATS, "raspergade/linDA_Order_Sig_LFC1.csv"),
    rasp_linDA_class      = file.path(INPUT_STATS, "raspergade/linDA_Class_Sig_LFC1.csv"),
    rasp_linDA_phylum     = file.path(INPUT_STATS, "raspergade/linDA_Phylum_Sig_LFC1.csv")
)

# ── Check which files exist ────────────────────────────────────────────────────
missing <- names(paths)[!sapply(paths, file.exists)]
if (length(missing) > 0) {
    message("⚠ Missing files (analysis will proceed with available data):")
    for (m in missing) message("  - ", m, " → ", paths[[m]])
}
```

```{r palettes}
# ── Shared aesthetics ──────────────────────────────────────────────────────────
PAL_BRANCH  <- c("Original" = "#00809D", "GCN-corrected" = "#9F1861")
PAL_VISIT   <- c("v2" = "#5CB338", "v6" = "#FFB200",
                 "v7" = "#A19AD3", "v8" = "#A31D1D")
THEME_BASE <-  theme_bw(base_size = 12)+
        theme(plot.title = element_text(hjust = 0.5,size=10,face = 'bold.italic'),
          plot.subtitle = element_text(hjust = 0.5,size=8),
              legend.position="none",
              axis.text.x=element_text(size=9,hjust=0.5,vjust = 0.5,angle=0),
              axis.text.y=element_text(size=9),
              axis.title.y = element_text(size=9),
              axis.title.x = element_text(size=9),
              strip.background = element_rect(fill = "#F5F5F5",color=NA,linewidth = 0),
              strip.text = element_text(face = "bold",size = 10,colour = "#2C3E50"),
              panel.grid.major.x = element_line(linewidth =0.3,linetype = "solid"),
              panel.grid.major.y = element_line(linewidth =0.2,linetype = "dashed", colour = "gray80"),
              panel.grid.minor.y = element_line(linewidth =0.2,linetype = "dotted", colour = "gray80"))

labels = c("v2"="Baseline 1", "v6"="Baseline 2", "v7"="1-week glasmacinal", "v8"="2-week glasmacinal")
labels_short = c("v2"="BL1", "v6"="BL2", "v7"="1W", "v8"="2W")
# 
# THEME_BASE  <- theme_bw(base_size = 12) +
#     theme(panel.grid.minor = element_blank(),
#           strip.background = element_rect(fill = "grey92"),
#           legend.position  = "bottom")

save_fig <- function(name, p, w = 10, h = 6) {
    ggsave(file.path(ANALYSIS_DIR,"figures", name), p, width = w, height = h, device = "pdf")
    invisible(p)
}
save_tbl <- function(name, tbl) {
    write_csv(tbl, file.path(ANALYSIS_DIR,"tables", name))
    invisible(tbl)
}
```

---

# 1  Feature Table Comparison

```{r load-tables}
# ── Load feature tables from QZA ──────────────────────────────────────────────
read_table_qza <- function(path, label) {
    if (!file.exists(path)) {
        warning("Table not found: ", path)
        return(NULL)
    }
    tbl <- read_qza(path)$data
    mat <- as(tbl, "matrix")
    tibble(
        branch        = label,
        total_reads   = sum(mat),
        n_asvs        = nrow(mat),
        n_samples     = ncol(mat),
        min_depth     = min(colSums(mat)),
        median_depth  = median(colSums(mat)),
        max_depth     = max(colSums(mat))
    )
}

orig_mat  <- if (file.exists(paths$orig_table))
    as(read_qza(paths$orig_table)$data, "matrix") else NULL

rasp_mat  <- if (file.exists(paths$rasp_table))
    as(read_qza(paths$rasp_table)$data, "matrix") else NULL

tbl_summary <- bind_rows(
    if (!is.null(orig_mat)) read_table_qza(paths$orig_table, "Original"),
    if (!is.null(rasp_mat)) read_table_qza(paths$rasp_table, "GCNs correction")
)

tbl_summary %>%
    mutate(across(where(is.numeric), ~ scales::comma(.x, accuracy = 1))) %>%
    kbl(caption = "Feature table summary: Original vs 16S Gene copy numbers (GCNs) correction ") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

save_tbl("01_feature_table_summary.csv", tbl_summary)
```

## 1.1 Per-sample sequencing depth

```{r depth-histogram}
library(patchwork)
if (!is.null(orig_mat) && !is.null(rasp_mat)) {
    depth_df <- bind_rows(
        tibble(branch = "Original",   sample = colnames(orig_mat), depth = colSums(orig_mat)),
        tibble(branch = "GCNs correction", sample = colnames(rasp_mat), depth = colSums(rasp_mat))
    ) %>%
        mutate(branch = factor(branch, levels = c("Original", "GCNs correction")))

    p_depth_hist <- ggplot(depth_df, aes(x = depth, fill = branch)) +
        geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
        scale_fill_manual(values = PAL_BRANCH) +
        scale_x_continuous(labels = comma) +
        facet_wrap(~branch, ncol = 1) +
        labs(#title = "Per-sample sequencing depth",
             x = "Reads per sample", y = "Number of samples", fill = NULL) +
        THEME_BASE + theme(legend.position = "none")

    print(p_depth_hist)
    save_fig("alpha/01_depth_histogram.pdf", p_depth_hist, w = 8, h = 6)

    # Paired per-sample depth comparison
    common_samples <- intersect(colnames(orig_mat), colnames(rasp_mat))
    depth_paired <- tibble(
        sample  = common_samples,
        orig    = colSums(orig_mat)[common_samples],
        rasp    = colSums(rasp_mat)[common_samples],
        pct_change = 100 * (rasp - orig) / orig
    )
    
    # correlations
  pearson_r <- cor(depth_paired$orig,depth_paired$rasp,method = "pearson",use = "complete.obs")
  spearman_rho <- cor(depth_paired$orig,depth_paired$rasp,method = "spearman",use = "complete.obs")

  lims <- range(c(depth_paired$orig, depth_paired$rasp), na.rm = TRUE)

    p_depth_scatter <- ggplot(depth_paired, aes(x = orig, y = rasp)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
      geom_smooth(method = "lm",se = FALSE,colour = "#F08D39",linewidth = 0.3) +
       geom_point(aes(colour = pct_change), size = 3, alpha = 0.8) +
        scale_colour_gradient2(low = "#2FA4D7", mid = "grey90", high = "#D6604D",
                               midpoint = 0, name = "% change") +
      coord_equal(xlim = lims, ylim = lims)+
      #xlim(lims)+ylim(lims)+
      scale_x_continuous(labels = comma) +
      scale_y_continuous(labels = comma) +
      labs(#title = "Per-sample depth: Original vs GCNs correction", subtitle = "Dashed line = no change; solid line = fitted trend",
             x = "Original reads", y = "16S GCN-corrected reads") +
      annotate("text",y =75e3,x = 135e3,
               hjust = 0,vjust = 1,label = paste0("Pearson r = ", round(pearson_r, 3),
                                                  "\nSpearman p = ", round(spearman_rho, 3)),
               size = 3) +
        THEME_BASE

    print(p_depth_scatter)
    save_fig("alpha/01_depth_scatter.pdf", p_depth_scatter, w = 7, h = 6)

    save_tbl("01_per_sample_depth.csv", depth_paired)
    
    p_combined <- p_depth_hist|p_depth_scatter
    p_combined <- p_combined + plot_annotation(tag_levels = 'A') + plot_layout(width = c(1,2))
    print(p_combined)
    save_fig("alpha/01_depth_combined.pdf",p_combined, w = 10, h = 6)
}
```

## 1.2 ASV abundance correlation

```{r asv-correlation}
if (!is.null(orig_mat) && !is.null(rasp_mat)) {
    common_asvs    <- intersect(rownames(orig_mat), rownames(rasp_mat))
    common_samples <- intersect(colnames(orig_mat), colnames(rasp_mat))

    orig_totals <- rowSums(orig_mat[common_asvs, common_samples])
    rasp_totals <- rowSums(rasp_mat[common_asvs, common_samples])

    asv_cor_df <- tibble(
        asv         = common_asvs,
        orig_abund  = orig_totals,
        rasp_abund  = rasp_totals,
        pct_change  = 100 * (rasp_totals - orig_totals) / (orig_totals + 1),
        log2_ratio  = log2((rasp_totals + 1) / (orig_totals + 1))
    ) %>%
        arrange(desc(abs(log2_ratio)))

    # Pearson + Spearman correlation
    r_pearson  <- cor(asv_cor_df$orig_abund, asv_cor_df$rasp_abund, method = "pearson")
    r_spearman <- cor(asv_cor_df$orig_abund, asv_cor_df$rasp_abund, method = "spearman")

    top_changed <- asv_cor_df %>% slice_head(n = 15)

    p_asv_cor <- ggplot(asv_cor_df, aes(x = log10(orig_abund + 1), y = log10(rasp_abund + 1))) +
        geom_point(aes(colour = log2_ratio), alpha = 0.5, size = 1.5) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
        scale_colour_gradient2(low = "#2166AC", mid = "grey80", high = "#D6604D",
                               midpoint = 0, name = "log2 ratio") +
        annotate("text", x = 0.2, y = max(log10(asv_cor_df$rasp_abund + 1)) * 0.95,
                 label = sprintf("Pearson r = %.3f\nSpearman ρ = %.3f", r_pearson, r_spearman),
                 hjust = 0, size = 3.5) +
        labs(#title = "ASV abundance correlation: Original vs RasperGade",
             x = "log10(Original total abundance + 1)",
             y = "log10(GCN-corrected total abundance + 1)") +
        THEME_BASE

    print(p_asv_cor)
    save_fig("alpha/01_asv_abundance_correlation.png", p_asv_cor, w = 8, h = 6)

    # Top 15 most-changed ASVs
    top_changed %>%
        mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
        kbl(caption = "Top 15 ASVs most affected by copy-number correction (by |log2 ratio|)") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

    save_tbl("01_asv_most_changed.csv", asv_cor_df)
}
```

## 1.3 Feature prevalence comparison

```{r prevalence}
if (!is.null(orig_mat) && !is.null(rasp_mat)) {
    common_asvs <- intersect(rownames(orig_mat), rownames(rasp_mat))

    prev_df <- tibble(
        asv          = common_asvs,
        prev_orig    = rowMeans(orig_mat[common_asvs, ] > 0),
        prev_rasp    = rowMeans(rasp_mat[common_asvs, ] > 0)
    )

    p_prev <- ggplot(prev_df, aes(x = prev_orig, y = prev_rasp)) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
        geom_point(alpha = 0.4, size = 1.5, colour = "#2166AC") +
        scale_x_continuous(labels = percent_format()) +
        scale_y_continuous(labels = percent_format()) +
        labs(title = "ASV prevalence: Original vs RasperGade",
             x = "Prevalence (Original)", y = "Prevalence (RasperGade)") +
        THEME_BASE

    print(p_prev)
    save_fig("alpha/01_prevalence.pdf", p_prev, w = 6, h = 6)
}
```

---

# 2  Alpha Diversity Comparison

```{r load-alpha}
# ── Load alpha diversity from QZA ─────────────────────────────────────────────
read_alpha_qza <- function(path, branch, metric) {
    if (!file.exists(path)) return(NULL)
  
    if(branch == 'Original' | metric== "Pielou Evenness"){
      df <- read_qza(path)$data %>%
          rownames_to_column("sample_id") %>%
          mutate(branch = branch, metric = metric)
      colnames(df)[2] <- "value"
    }else{
      df <- read_qza(path)$data%>%
        select(sample_id=V1,value=V2) %>%
        mutate(branch = branch, metric = metric)
    }
  df
    
}

meta <- if (file.exists(paths$metadata))
    read_tsv(paths$metadata, show_col_types = FALSE) %>%
        mutate(visit = case_when(
                      timepoint == "Baseline 1"   ~ "v2",
                      timepoint == "Baseline 2"   ~ "v6",
                      timepoint == "1-week EP395" ~ "v7",
                      timepoint == "2-week EP395" ~ "v8",
                      TRUE                        ~ timepoint   # keep everything else the same
                    )) %>%
        rename(sample_id = `sample-id`) else NULL
visits_levels <- c("v2", "v6", "v7", "v8")

alpha_df <- bind_rows(
    read_alpha_qza(paths$orig_faith,    "Original",   "Faith
                   PD"),
    read_alpha_qza(paths$rasp_faith,    "GCN-corrected", "Faith PD"),
    read_alpha_qza(paths$orig_evenness, "Original",   "Pielou Evenness"),
    read_alpha_qza(paths$rasp_evenness, "GCN-corrected", "Pielou Evenness")
) %>%
    mutate(branch = factor(branch, levels = c("Original", "GCN-corrected")))

# Merge with metadata
if (!is.null(meta)) {
    alpha_df <- alpha_df %>%
        left_join(meta, by = "sample_id") %>%
      mutate(visit = factor(visit, levels = visits_levels))
}

cat("Alpha diversity data loaded:", nrow(alpha_df), "observations\n")
```

## 2.1 Side-by-side boxplots

```{r alpha-boxplots, fig.width=12, fig.height=6}
if (nrow(alpha_df) > 0) {
    p_alpha_box <- ggplot(alpha_df, aes(x = branch, y = value, fill = branch)) +
        geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.6) +
        geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
        scale_fill_manual(values = PAL_BRANCH) +
        
        facet_wrap(.~metric, scales = "free_y", ncol = 1) +
        labs(#title = "Alpha diversity: Original vs RasperGade",
             x = NULL, y = "Diversity value", fill = NULL) +
        THEME_BASE + theme(legend.position = "none")

    print(p_alpha_box)
    save_fig("alpha/02_alpha_boxplots.pdf", p_alpha_box, w = 10, h = 5)
}
```

## 2.2 Timepoint-stratified boxplots

```{r alpha-by-timepoint, fig.width=14, fig.height=6}
if (nrow(alpha_df) > 0 && "timepoint" %in% colnames(alpha_df)) {
    p_alpha_tp <- ggplot(alpha_df,
                         aes(x = visit, y = value, fill = branch)) +
        geom_boxplot(outlier.shape = NA, alpha = 0.7,
                     position = position_dodge(0.8), width = 0.7) +
        geom_point(position = position_dodge(0.8),
                   alpha = 0.5, size = 1.5) +
        scale_fill_manual(values = PAL_BRANCH) +
        scale_x_discrete(labels=labels_short) +
        facet_wrap(~metric, scales = "free_y", ncol =1) +
        labs(#title = "Alpha diversity by timepoint: Original vs RasperGade",
             x = "Timepoint", y = "Diversity value", fill = "Branch") +
        THEME_BASE

    print(p_alpha_tp)
    save_fig("alpha/02_alpha_by_timepoint.pdf", p_alpha_tp, w = 12, h = 6)
    
    p_combined <-     p_alpha_box | p_alpha_tp
    p_combined <- p_combined + plot_annotation(tag_levels = 'A') + plot_layout(width = c(1,2))
    print(p_combined)
    save_fig("alpha/02_alpha_combined.pdf",p_combined, w = 8, h = 8)
}
```

## 2.3 Paired per-sample comparison

```{r alpha-paired, fig.width=10, fig.height=5}
if (nrow(alpha_df) > 0) {
    alpha_wide <- alpha_df %>%
        select(sample_id, branch, metric, value) %>%
        pivot_wider(names_from = branch, values_from = value)

    if ("Original" %in% colnames(alpha_wide) &&  "GCNs correction" %in% colnames(alpha_wide)) {
        p_paired <- ggplot(alpha_wide, aes(x = Original, y = `GCNs correction`)) +
            geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
            geom_point(alpha = 0.6, size = 2, colour = "#2166AC") +
            facet_wrap(~metric, scales = "free", ncol = 2) +
            labs(title = "Per-sample alpha diversity: Original vs RasperGade",
                 subtitle = "Dashed line = no change") +
            THEME_BASE

        print(p_paired)
        save_fig("alpha/02_alpha_paired.png", p_paired, w = 10, h = 5)

        # Correlation statistics
        alpha_cor <- alpha_wide %>%
            group_by(metric) %>%
            summarise(
                pearson_r  = cor(Original, `GCNs correction`, method = "pearson",  use = "complete.obs"),
                spearman_r = cor(Original, `GCNs correction`, method = "spearman", use = "complete.obs"),
                mean_diff  = mean(`GCNs correction` - Original, na.rm = TRUE),
                sd_diff    = sd(`GCNs correction` - Original, na.rm = TRUE)
            )

        alpha_cor %>%
            mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
            kbl(caption = "Alpha diversity correlation: Original vs RasperGade") %>%
            kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

        save_tbl("02_alpha_diversity_correlation.csv", alpha_cor)
    }
}
```

## 2.4 LMM statistics comparison

```{r lmm-table}
read_csv_safe <- function(path) {
    if (file.exists(path)) read_csv(path, show_col_types = FALSE) else NULL
}

make_lmm_comparison <- function(metric_name, orig_anova_path, rasp_anova_path) {
    o <- read_csv_safe(orig_anova_path)
    r <- read_csv_safe(rasp_anova_path)
    if (is.null(o) || is.null(r)) return(NULL)

    bind_rows(
        mutate(o, branch = "Original",   metric = metric_name),
        mutate(r, branch = "RasperGade", metric = metric_name)
    )
}

lmm_faith <- make_lmm_comparison(
    "Faith PD",
    paths$orig_faith_anova,
    paths$rasp_faith_anova
)

lmm_even <- make_lmm_comparison(
    "Pielou Evenness",
    paths$orig_even_anova,
    paths$rasp_even_anova
)

lmm_all <- bind_rows(lmm_faith, lmm_even)

if (nrow(lmm_all) > 0) {
    lmm_all %>%
        mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
        arrange(metric, branch) %>%
        kbl(caption = "LMM ANOVA statistics: Original vs RasperGade") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
        row_spec(which(lmm_all$`Pr(>F)` < 0.05), bold = TRUE, color = "white", background = "#2166AC")

    save_tbl("02_lmm_anova_comparison.csv", lmm_all)
}
```

## 2.5 Pairwise contrasts comparison

```{r emmeans-table}
make_emmeans_comparison <- function(metric_name, orig_path, rasp_path) {
    o <- read_csv_safe(orig_path)
    r <- read_csv_safe(rasp_path)
    if (is.null(o) || is.null(r)) return(NULL)

    full_join(o, r, by = "contrast", suffix = c("_orig", "_rasp")) %>%
        mutate(
            metric = metric_name,
            sig_orig = p.value_orig < 0.05,
            sig_rasp = p.value_rasp < 0.05,
            sig_changed = sig_orig != sig_rasp
        ) %>%
        arrange(contrast)
}

emmeans_faith <- make_emmeans_comparison(
    "Faith PD",
    paths$orig_faith_emmeans,
    paths$rasp_faith_emmeans
)

emmeans_even <- make_emmeans_comparison(
    "Pielou Evenness",
    paths$orig_even_emmeans,
    paths$rasp_even_emmeans
)

emmeans_all <- bind_rows(emmeans_faith, emmeans_even)

if (!is.null(emmeans_all) && nrow(emmeans_all) > 0) {
    emmeans_all %>%
        select(metric, contrast, estimate_orig, p.value_orig, estimate_rasp, p.value_rasp,
               sig_orig, sig_rasp, sig_changed) %>%
        mutate(across(c(estimate_orig, estimate_rasp, p.value_orig, p.value_rasp),
                      ~ round(.x, 4))) %>%
        kbl(caption = "Pairwise contrasts: Original vs RasperGade") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
        row_spec(which(emmeans_all$sig_changed), bold = TRUE, background = "#FFF3CD")

    n_changed <- sum(emmeans_all$sig_changed, na.rm = TRUE)
    cat(sprintf("\n⚠ Contrasts that changed significance: %d / %d\n",
                n_changed, nrow(emmeans_all)))

    save_tbl("02_emmeans_comparison.csv", emmeans_all)
}
```

---

# 3  Beta Diversity Comparison

```{r load-beta}
read_dist_qza <- function(path) {
    if (!file.exists(path)) return(NULL)
    as.dist(read_qza(path)$data)
}

dist_orig <- read_dist_qza(paths$orig_wunifrac)
dist_rasp <- read_dist_qza(paths$rasp_wunifrac)
```

```{r}

plot_beta_diversity <- function(pcoaResults,metadata,title){
  library(qiime2R)
  library(ggplot2)
  library(dplyr)
  library(tidyverse)
  library(ggrepel)
  library(ggforce)
  library(ggtext)
  library(ggside)

  pca_tbl<- pcoaResults$data$Vectors
  
  pc_variances <- pca_tbl %>%  summarize(across(everything(), var))
  total_variance <- sum(pc_variances,na.rm = T)
  proportion_variance <- round(100.0 * pc_variances / total_variance,2)  

  pca_tbl<-merge(x=pca_tbl %>% select(sample_id=SampleID,PC1,PC2),y=metadata,by='sample_id')
  
  pca_tbl$ovisit <- pca_tbl$visit
  pca_tbl <- pca_tbl %>%
    mutate(visit = case_when(
      ovisit == "v2" ~ "BL1",
      ovisit == "v6" ~ "BL2",
      ovisit == "v7" ~ "1W",
      ovisit == "v8" ~ "2W",
      TRUE ~ ovisit 
    )) %>% mutate(visit=factor(visit,levels=c("BL1", "BL2",
                                              "1W", "2W")))
   print(str(pca_tbl[, c("visit", "PC1", "PC2")]))
 
   p<- ggplot(data=pca_tbl,aes(shape=visit,x=PC2, y=PC1)) +
          geom_point(aes(fill=visit),size=3,alpha=.7,stroke=0.5,color='black')+
          scale_shape_manual(name="",values=c(21,24,22,23)) +
          scale_fill_manual(name="",values =c("#5CB338", "#FFB200","#A19AD3", "#A31D1D") )+
          labs(y=paste0("PCoA1 [",proportion_variance$PC1,"%]"),x=paste0("PCoA2 [",proportion_variance$PC2,"%]"))+
          annotate("text", x = -0.29, y = 0.34,
                 label = title,
                 hjust = 0, size = 3.5) +
          theme_bw()+
          theme(text=element_text(size=10),
                axis.title=element_text(size=10),
                legend.text=element_text(size=9),
                legend.title = element_text(size=10),
                legend.position="bottom", legend.box="horizontal",
                axis.title.x = element_text(size = 9, margin = margin(t = 5, unit = "pt")),
                axis.title.y = element_text(size = 9, margin = margin(r = 0, unit = "pt"))
                ) +
    # Y-side: Boxplots of PC1 ~ visit (vertical orientation, shared y-axis)
    geom_ysideboxplot(aes(x = visit, y = PC1, fill = visit),  orientation = "x",# Local aes: use PC1 on shared y
                      inherit.aes = FALSE, varwidth = TRUE, alpha = 0.7,linewidth = 0.2, show.legend = FALSE) +
       theme(ggside.axis.text.y = element_blank(),
          ggside.axis.ticks.y =  element_blank())+



# X-side: Boxplots of PC2 ~ visit (horizontal orientation, shared x-axis)
geom_xsideboxplot(aes(y = visit, x = PC2, fill = visit), orientation = "y", # Local aes: use PC2 on shared x
                  inherit.aes = FALSE, varwidth = TRUE, alpha = 0.7,linewidth = 0.2, show.legend = FALSE) +
#theme(ggside.axis.text.x = element_text(size = 10, angle = -90, hjust = 0, vjust = 0)) +
theme(ggside.axis.text.x = element_blank(),
      ggside.axis.ticks.x= element_blank())+
    guides(shape = guide_legend(ncol = 4)) + 
theme(legend.position = "bottom",
      legend.justification = "center",
      legend.margin = margin(0, 0, 0, 0),  # remove margins
      legend.box.margin = margin(0, 0, 0, 0),  # negative top margin to pull closer
      legend.direction = "horizontal")  # horizontal legend often works better at bottom
  
  return(p)
}


```

## 3.1 PCoA ordinations

```{r pcoa-plots}
make_pcoa_plot <- function(dist, title, meta = NULL, color_col = "visit") {
    if (is.null(dist)) return(NULL)
    pcoa_res <- cmdscale(dist, k = 2, eig = TRUE)
    var_exp  <- pcoa_res$eig / sum(abs(pcoa_res$eig)) * 100

    df <- as_tibble(pcoa_res$points, .name_repair = "unique") %>%
        setNames(c("PC1", "PC2")) %>%
        mutate(sample_id = labels(dist))

    if (!is.null(meta) && color_col %in% colnames(meta)) {
        df <- left_join(df, meta %>% select(sample_id, all_of(color_col)), by = "sample_id")
    } else {
        df[[color_col]] <- NA
    }

    ggplot(df, aes(x = PC1, y = PC2, colour = .data[[color_col]])) +
        geom_point(size = 3, alpha = 0.8) +
        stat_ellipse(level = 0.8, linetype = "dashed", linewidth = 0.5) +
        scale_colour_manual(values = PAL_VISIT, na.value = "grey50",
                            name = color_col) +
        labs(title = title,
             x = sprintf("PC1 (%.1f%%)", var_exp[1]),
             y = sprintf("PC2 (%.1f%%)", var_exp[2])) +
        THEME_BASE
}

#p_pcoa_orig <- make_pcoa_plot(dist_orig, "Weighted UniFrac PCoA — Original", meta)
#p_pcoa_rasp <- make_pcoa_plot(dist_rasp, "Weighted UniFrac PCoA — RasperGade", meta)

p_pcoa_orig<- plot_beta_diversity(read_qza(paths$orig_wunifrac_pcoa),meta,"Original")
p_pcoa_rasp<- plot_beta_diversity(read_qza(paths$rasp_wunifrac_pcoa),meta,"16S GCN-corrected")

if (!is.null(p_pcoa_orig) && !is.null(p_pcoa_rasp)) {
    p_pcoa_combined <- p_pcoa_orig/p_pcoa_rasp +
        plot_layout(guides = "collect") &
        theme(legend.position = "top")
    print(p_pcoa_combined)
    save_fig("beta/03_pcoa_comparison.pdf", p_pcoa_combined, w = 6, h = 10)
}
```

## 3.2 Procrustes analysis

```{r procrustes}
if (!is.null(dist_orig) && !is.null(dist_rasp)) {
  
    labels = list("Original" = "Original", "RasperGade"="16S GCN-corrected")
    common_samples <- intersect(labels(dist_orig), labels(dist_rasp))

    mat_orig <- as.matrix(dist_orig)[common_samples, common_samples]
    mat_rasp <- as.matrix(dist_rasp)[common_samples, common_samples]

    pcoa_orig <- cmdscale(as.dist(mat_orig), k = min(length(common_samples) - 1, 10))
    pcoa_rasp <- cmdscale(as.dist(mat_rasp), k = min(length(common_samples) - 1, 10))


    proc <- procrustes(pcoa_orig, pcoa_rasp, symmetric = TRUE)
    prot <- protest(pcoa_orig, pcoa_rasp, permutations = 9999)

    cat(sprintf("Procrustes M² = %.4f\n", proc$ss))
    cat(sprintf("Procrustes r  = %.4f\n", sqrt(1 - proc$ss)))
    cat(sprintf("Protest p     = %.4f  (9999 permutations)\n", prot$signif))

    # Procrustes plot
    proc_df <- data.frame(
        orig_x = pcoa_orig[, 1], orig_y = pcoa_orig[, 2],
        rasp_x = proc$Yrot[, 1], rasp_y = proc$Yrot[, 2],
        sample  = common_samples
    )

    p_proc <- ggplot(proc_df) +
        geom_segment(aes(x = orig_x, y = orig_y, xend = rasp_x, yend = rasp_y),
                     arrow = arrow(length = unit(0.15, "cm")),
                     alpha = 0.5, colour = "grey50") +
        geom_point(aes(x = orig_x, y = orig_y,colour = "Original",shape = "Original"),
                   size = 2.5, alpha = 0.8) +
        geom_point(aes(x = rasp_x, y = rasp_y,colour = "RasperGade", shape = "RasperGade"),
                   size = 2.5, alpha = 0.8) +
      
      scale_colour_manual(values = c("Original" = "#00809D","RasperGade" = "#9F1861"),labels = labels) +
      scale_shape_manual(values = c("Original" = 16,"RasperGade" = 17),labels = labels) +
      
      annotate("text", x = min(c(proc_df$orig_x, proc_df$rasp_x)),
                 y = max(c(proc_df$orig_y, proc_df$rasp_y)),
                 label = sprintf("M² = %.4f\nr = %.4f\np = %.4f",
                                 proc$ss, sqrt(1 - proc$ss), prot$signif),
                 hjust = 0, vjust = 1, size = 3.5) +
        labs(#title = "Procrustes analysis: Original (●) vs RasperGade (▲)",
            #subtitle = "Arrows connect the same sample between ordinations",
             x = "PC1", y = "PC2") +
   
        THEME_BASE + theme(legend.position = c(0.7, 0.05),
                          legend.box = "horizontal",
                          legend.text = element_text(size=9),
                          legend.background = element_rect(fill = NA, colour = NA))+
       guides(shape = guide_legend(title = NULL,override.aes = list(size = 3,color=c("#00809D","#9F1861")),nrow = 1),
              colour = "none")  

    print(p_proc)
    save_fig("beta/03_procrustes.pdf", p_proc, w = 7, h = 7)

    # Mantel test
    mantel_res <- mantel(dist_orig, dist_rasp, method = "pearson", permutations = 9999)
    cat(sprintf("\nMantel r = %.4f, p = %.4f\n", mantel_res$statistic, mantel_res$signif))

    proc_summary <- tibble(
        test     = c("Procrustes M²", "Procrustes r", "Protest p",
                     "Mantel r", "Mantel p"),
        value    = round(c(proc$ss, sqrt(1 - proc$ss), prot$signif,
                           mantel_res$statistic, mantel_res$signif), 4),
        interpretation = c("0 = identical; 1 = unrelated",
                           "1 = identical ordinations",
                           "< 0.05 = ordinations differ significantly",
                           "1 = identical distances",
                           "< 0.05 = distance matrices differ significantly")
    )

    proc_summary %>%
        kbl(caption = "Procrustes + Mantel comparison") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

    save_tbl("03_procrustes_mantel.csv", proc_summary)
}
```

## 3.3 PERMANOVA comparison

```{r permanova-table}
perm_orig <- read_csv_safe(paths$orig_perm)
perm_rasp <- read_csv_safe(paths$rasp_perm)

normalize_pairs <- function(pair_string) {
  parts <- str_split(pair_string, " vs ")[[1]]
  parts_sorted <- sort(parts)
  paste(parts_sorted, collapse = " vs ")
}


if (!is.null(perm_orig) && !is.null(perm_rasp)) {
  
  
     perm_combined <- full_join(
        perm_orig %>% 
          mutate(pairs_norm = sapply(pairs, normalize_pairs)) %>%
          select(pairs_norm, R2_orig = R2, p_orig = p.value),
        perm_rasp %>% 
          mutate(pairs_norm = sapply(pairs, normalize_pairs)) %>%
          select(pairs_norm, R2_rasp = R2, p_rasp = p.value),
        by = "pairs_norm"
      ) %>%
        rename(pairs = pairs_norm) 
     
   perm_combined <- perm_combined %>%
        mutate(
            sig_orig    = p_orig < 0.05,
            sig_rasp    = p_rasp < 0.05,
            sig_changed = sig_orig != sig_rasp,
            R2_delta    = R2_rasp - R2_orig
        )

    perm_combined %>%
        mutate(across(where(is.numeric), ~ round(.x, 4)),
               sig_changed = ifelse(sig_changed, "⚠ YES", "No")) %>%
        kbl(caption = "PERMANOVA R² and p-values: Original vs RasperGade") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
        row_spec(which(perm_combined$sig_changed), bold = TRUE, background = "#FFF3CD")

    save_tbl("03_permanova_comparison.csv", perm_combined)
    perm_combined <- perm_combined %>%
      mutate(
      pairs_new = case_when(
        pairs == "v6 vs v8" ~ "BL2 vs 2W",
        pairs == "v6 vs v7" ~ "BL2 vs 1W",
        pairs == "v2 vs v6" ~ "BL2 vs BL1",
        pairs == "v7 vs v8" ~ "1W vs 2W",
        pairs == "v2 vs v8" ~ "BL1 vs 2W",
        pairs == "v2 vs v7" ~ "BL1 vs 1W",
        pairs == "v7 vs v8" ~ "1W vs 2W",
        pairs == "v2 vs v8" ~ "BL1 vs 2W",
        TRUE ~ pairs   # keep original if unmatched
      ),
      sig_stat=case_when(
        sig_orig & sig_rasp ~ "Both significant",
        !sig_orig & !sig_rasp ~ "Neither significant",
        sig_orig & !sig_rasp ~ "Only original",
        !sig_orig & sig_rasp ~ "Only 16S GCN-corrected",
        TRUE ~ "Other"
      )
    )
      
    # R² comparison plot
    p_r2 <- ggplot(perm_combined, aes(x = R2_orig, y = R2_rasp, label = pairs_new)) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
        geom_point(aes(fill = sig_stat), shape=21,size = 3,alpha=0.7,color="black",stroke=0.3) +
        geom_text_repel(size = 3) +
        scale_fill_manual(values = c("Both significant" = "#A8D966", 
                                       "Neither significant" = "#1591DC",
                                       "Only original"="#D6336C",
                                       "Only 16S GCN-corrected"="#FF9D23",
                                       "other"="#BBD5DA"),
                            name = "") +
        labs(#title = "PERMANOVA R² comparison",
             x = "R² - Original", y = "R² - 16S GCN-corrected") +
        THEME_BASE + theme(legend.position = c(0.23, 0.9),
                          legend.text = element_text(size=9),
                          legend.background = element_rect(fill = NA, colour = NA))+
       guides(fill = guide_legend(title = NULL,override.aes = list(size = 3),ncol = 1))  

    print(p_r2)
    save_fig("beta/03_permanova_r2.pdf", p_r2, w = 7, h = 6)
}
```

```{r}
p_pcoa <- (p_pcoa_orig + p_pcoa_rasp) + plot_layout(guides = "collect") &
        theme(legend.position = "top")
p_beta_combined <-  p_pcoa / (p_r2 + p_proc) + plot_annotation(tag_levels = "A")
save_fig("beta/03_beta-div_combined.pdf", p_beta_combined, w = 10, h = 10)
```

---

# 4  Differential Abundance Comparison

```{r load-da}
# ── Load all LinDA results ─────────────────────────────────────────────────────
levels <- c("ASV", "Genus", "Family", "Order", "Class", "Phylum")

da_paths <- list(
    ASV    = list(orig = paths$orig_linDA_asv,    rasp = paths$rasp_linDA_asv),
    Genus  = list(orig = paths$orig_linDA_genus,  rasp = paths$rasp_linDA_genus),
    Family = list(orig = paths$orig_linDA_family, rasp = paths$rasp_linDA_family),
    Order  = list(orig = paths$orig_linDA_order,  rasp = paths$rasp_linDA_order),
    Class  = list(orig = paths$orig_linDA_class,  rasp = paths$rasp_linDA_class),
    Phylum = list(orig = paths$orig_linDA_phylum, rasp = paths$rasp_linDA_phylum)
)

load_da <- function(path, branch, level) {
    df <- read_csv_safe(path)
    if (is.null(df)) return(NULL)
    df %>% mutate(branch = branch, level = level)
}

da_all <- map_dfr(levels, function(lv) {
    bind_rows(
        load_da(da_paths[[lv]]$orig, "Original",   lv),
        load_da(da_paths[[lv]]$rasp, "RasperGade", lv)
    )
})

if (nrow(da_all) > 0) {
    cat("Differential abundance data loaded.\n")
    cat("Levels available:", paste(unique(da_all$level), collapse = ", "), "\n")
    cat("Branches available:", paste(unique(da_all$branch), collapse = ", "), "\n")
}
```

## 4.1 Number of significant taxa per level

```{r da-counts, fig.width=10, fig.height=5}
if (nrow(da_all) > 0) {
    da_counts <- da_all %>%
        group_by(level, branch, visit) %>%
        summarise(n_sig = n(), .groups = "drop") %>%
        mutate(level = factor(level, levels = levels))

    p_da_counts <- ggplot(da_counts, aes(x = level, y = n_sig, fill = branch)) +
        geom_col(position = position_dodge(0.75), width = 0.7) +
        scale_fill_manual(values = PAL_BRANCH) +
        facet_wrap(~visit) +
        labs(title = "Number of significant taxa per level and visit",
             x = "Taxonomic level", y = "Significant taxa (|LFC| > 1, padj < 0.05)",
             fill = "Branch") +
        THEME_BASE

    print(p_da_counts)
    save_fig("da/04_sig_taxa_counts.pdf", p_da_counts, w = 12, h = 5)

    da_counts_wide <- da_counts %>%
        pivot_wider(names_from = branch, values_from = n_sig, values_fill = 0) %>%
        mutate(delta = RasperGade - Original)

    da_counts_wide %>%
        kbl(caption = "Significant taxa count: Original vs RasperGade") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

    save_tbl("04_sig_taxa_counts.csv", da_counts_wide)
}
```

## 4.2 LFC correlation between branches

```{r lfc-correlation, fig.width=14, fig.height=10}
if (nrow(da_all) > 0) {
    # Join by taxa + visit
    taxa_col <- if ("taxa" %in% colnames(da_all)) "taxa" else "Taxon"

    da_wide <- da_all %>%
        select(all_of(taxa_col), visit, branch, level, log2FoldChange, padj) %>%
        pivot_wider(
            names_from  = branch,
            values_from = c(log2FoldChange, padj),
            names_glue  = "{.value}_{branch}"
        )

    p_lfc_cor <- ggplot(da_wide,
                        aes(x = log2FoldChange_Original,
                            y = log2FoldChange_RasperGade)) +
        geom_hline(yintercept = 0, linetype = "dashed", colour = "grey70") +
        geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
        geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = "dotted") +
        geom_point(alpha = 0.5, size = 1.8, colour = "#2166AC") +
        facet_grid(visit ~ level) +
        labs(title = "Log2 fold change correlation: Original vs RasperGade",
             x = "LFC — Original", y = "LFC — RasperGade") +
        THEME_BASE

    print(p_lfc_cor)
    save_fig("da/04_lfc_correlation.pdf", p_lfc_cor, w = 14, h = 10)

    # Correlation stats per level
    lfc_cor_stats <- da_wide %>%
        group_by(level) %>%
        summarise(
            n        = n(),
            pearson  = cor(log2FoldChange_Original, log2FoldChange_RasperGade,
                           use = "complete.obs", method = "pearson"),
            spearman = cor(log2FoldChange_Original, log2FoldChange_RasperGade,
                           use = "complete.obs", method = "spearman"),
            .groups  = "drop"
        ) %>%
        mutate(across(c(pearson, spearman), ~ round(.x, 4)))

    lfc_cor_stats %>%
        kbl(caption = "LFC Pearson and Spearman correlation per taxonomic level") %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

    save_tbl("04_lfc_correlation.csv", lfc_cor_stats)
}
```

## 4.3 Overlap between significant taxa (UpSet plot)

```{r upset-plots, fig.width=10, fig.height=6}
if (nrow(da_all) > 0) {
    taxa_col <- if ("taxa" %in% colnames(da_all)) "taxa" else "Taxon"

    make_upset <- function(da_df, level_filter, title_sfx) {
        df_filt <- da_df %>%
            filter(level == level_filter)

        if (nrow(df_filt) == 0) return(invisible(NULL))

        taxa_orig <- unique(df_filt$taxa[df_filt$branch == "Original"])
        taxa_rasp <- unique(df_filt$taxa[df_filt$branch == "RasperGade"])

        upset_list <- list(
            Original   = taxa_orig,
            RasperGade = taxa_rasp
        )

        shared    <- length(intersect(taxa_orig, taxa_rasp))
        only_orig <- length(setdiff(taxa_orig, taxa_rasp))
        only_rasp <- length(setdiff(taxa_rasp, taxa_orig))

        cat(sprintf("%s — Shared: %d | Only Original: %d | Only RasperGade: %d\n",
                    level_filter, shared, only_orig, only_rasp))

        upset(fromList(upset_list),
              sets.bar.color  = unname(PAL_BRANCH),
              main.bar.color  = "grey30",
              text.scale      = 1.3,
              order.by        = "freq")

        title(main = sprintf("Overlap of significant %s — %s", level_filter, title_sfx),
              line = 2.5)

        tibble(level = level_filter,
               shared = shared, only_original = only_orig, only_raspergade = only_rasp,
               jaccard = shared / (shared + only_orig + only_rasp))
    }

    overlap_summary <- map_dfr(levels, function(lv) {
        tryCatch({
            png(file.path("figures/da", sprintf("04_upset_%s.png", tolower(lv))),
                width = 900, height = 600, res = 100)
            res <- make_upset(da_all, lv, "Original vs RasperGade")
            dev.off()
            res
        }, error = function(e) {
            dev.off()
            NULL
        })
    })

    if (!is.null(overlap_summary) && nrow(overlap_summary) > 0) {
        overlap_summary %>%
            mutate(jaccard = round(jaccard, 3)) %>%
            kbl(caption = "Overlap of significant taxa: Original vs RasperGade") %>%
            kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

        save_tbl("04_overlap_summary.csv", overlap_summary)
    }
}
```

## 4.4 Taxa gained and lost after correction

```{r gained-lost}
if (nrow(da_all) > 0) {
    taxa_col <- if ("taxa" %in% colnames(da_all)) "taxa" else "Taxon"

    gained_lost <- map_dfr(levels, function(lv) {
        df_lv     <- da_all %>% filter(level == lv)
        orig_taxa <- unique(df_lv[[taxa_col]][df_lv$branch == "Original"])
        rasp_taxa <- unique(df_lv[[taxa_col]][df_lv$branch == "RasperGade"])

        bind_rows(
            tibble(level = lv, status = "Only in Original",
                   taxa = setdiff(orig_taxa, rasp_taxa)),
            tibble(level = lv, status = "Only in RasperGade",
                   taxa = setdiff(rasp_taxa, orig_taxa))
        )
    })

    if (nrow(gained_lost) > 0) {
        gained_lost %>%
            group_by(level, status) %>%
            summarise(n = n(), .groups = "drop") %>%
            pivot_wider(names_from = status, values_from = n, values_fill = 0) %>%
            kbl(caption = "Taxa gained/lost at each taxonomic level") %>%
            kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

        save_tbl("04_taxa_gained_lost.csv", gained_lost)
    }
}
```

## 4.5 Heatmap of shared significant taxa (Genus level)

```{r da-heatmap, fig.width=12, fig.height=8}
if (nrow(da_all) > 0) {
    da_genus <- da_all %>% filter(level == "Genus")

    if (nrow(da_genus) > 0) {
        taxa_col <- if ("taxa" %in% colnames(da_genus)) "taxa" else "Taxon"

        shared_genera <- intersect(
            unique(da_genus[[taxa_col]][da_genus$branch == "Original"]),
            unique(da_genus[[taxa_col]][da_genus$branch == "RasperGade"])
        )

        if (length(shared_genera) > 0) {
            heat_df <- da_genus %>%
                filter(.data[[taxa_col]] %in% shared_genera) %>%
                mutate(label = paste(visit, branch, sep = " | ")) %>%
                select(all_of(taxa_col), label, log2FoldChange) %>%
                pivot_wider(names_from = label, values_from = log2FoldChange) %>%
                column_to_rownames(taxa_col)

            lfc_range <- max(abs(heat_df), na.rm = TRUE)
            
            heat_mat <- as.matrix(heat_df)
            heat_mat[is.na(heat_mat)] <- 0
            
            pheatmap(
                heat_mat,
                color    = colorRampPalette(c("#2166AC", "white", "#D6604D"))(51),
                breaks   = seq(-lfc_range, lfc_range, length.out = 52),
                cluster_rows = TRUE,
                cluster_cols = FALSE,
                main     = "LFC of shared significant genera: Original vs RasperGade",
                fontsize = 9,
                filename = file.path(ANALYSIS_DIR,"figures/da/04_heatmap_genus.pdf"),
                width    = 12, height = 8
            )

            pheatmap(
                heat_mat,
                color    = colorRampPalette(c("#2166AC", "white", "#D6604D"))(51),
                breaks   = seq(-lfc_range, lfc_range, length.out = 52),
                cluster_rows = TRUE,
                cluster_cols = FALSE,
                main     = "LFC of shared significant genera: Original vs RasperGade",
                fontsize = 9
            )
        }
    }
}
```

# Session Info

```{r session-info}
sessionInfo()
```
